Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
completed one editing pass through the MS, and got below the page limit

@@

In two major areas, our lack of skills or knowledge resulted in ugly
Haskell.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansome-peyton-jones},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, we found GHC's call-site and allocation
profiles did not provide enough information to be useful.
We~miss valgrind, whose callgrind profiler gives us enough information
without \emph{any} annotations in the source.

@@

\section{Our previous code base compared}

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{Menke:2008wu}.

Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
data structures such as mutable oct-trees, vectors, and arrays, with 
significant amounts of clever pointer arithmetic.
Our enhancement involved 
modifying Matt to incorporate sequence information along with the structural 
aligner.
While we initially estimated this would be a three-month project, it 
took most of a year because the mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: since 
invariants were not clear, nearly every change resulted in new segfaults.
One 
enhancement, partial alignments (essentially a cosmetic manipulation of the 
output, but requiring deep information about some of the data structures) was 
deemed infeasible.
We believe we could have completed a ground-up rewrite in 
Haskell, with most of the run-time performance, in fewer than nine months.

In general, the failure modes we encounter most in C++ programming are 
segmentation faults, memory errors (due to pointer arithmetic and uninitialized 
memory) and heap exhaustion.
In contrast, the failure modes we most commonly 
see in Haskell code are type-check failures, or at worst, runtime errors around 
bounds checking of arrays and vectors.
While bounds checking failures have 
proved difficult to debug, they are still less so than C++ memory errors.

Haskell makes it difficult to write code in haste, because Haskell militates 
towards careful type-checking and clearly thought-out function contracts.

Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
programmer who attempts to write code quickly with the goal of later making it 
work correctly.
We have found MRFy to be easier to enhance and maintain than 
our existing C++ code bases.
Despite the learning curve of Haskell, we 
implemented most of MRFy in roughly three months of part-time work.

\section{State of the practice}

\subsection{Our group then}
Programmers generally avoid reimplementing algorithms, but there may be good reasons for doing so.
Although the HMMER software contains an mature C++ implementation of the Viterbi algorithm,
a previous member of our group implemented \textit{multidimensional} dynamic programming for SMURF
in haste, resulting in 12,000 lines of undocumented code (including one function of 700 lines).
Though we are using the same Viterbi algorithm HMMER implements, we chose to reimplement it because
we wished to explore the design space of stochastic search techniques using higher order functions
in Haskell.
We found that reimplementing the Viterbi algorithm in Haskell was easy and pleasant.
In contrast, the programming effort to transform SMURF into SMURFLite, a minor evolution, required
several months of research to simply understand the SMURF codebase.

A set of enhancements to another large C++ implementation, the Matt~\cite{Menke:2008wu} structural
aligner, required nearly a year of programming due to complex mutable data structures being 
hard to repurpose.
In fact, we deemed one enhancement infeasible.
Our experience working with these C++ codebases largely involved the failure modes of
segmentation faults, memory errors, and heap exhaustion.

\subsection{Our group now}
Implementing MRFy in Haskell allowed us great flexibility to explore the space of stochastic search
techniques, initial guess heuristics, and search termination.
Higher order functions militate towards more flexibility, so that adding several planned features 
to MRFy should be easy.
The first functioning version of MRFy took roughly three months of part-time work.
The Haskell failure modes we have experienced have largely been typecheck errors; at the worst,
we have seen runtime errors around bounds checking of arrays and vectors, though we wanted
better debugging tools for these runtime errors.

\subsection{Computational biology at large}
There exists some excellent software in the computational biology community, though little of it
has been written in Haskell.
The training component of MRFy was modified from that of HMMER, and working with the HMMER
codebase was relatively pleasant, as data structures and their mutators are well documented.
BioHaskell exists (and we use \texttt{Bio.Sequence.Fasta} as a parser) but is not nearly as
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that BioHaskell and other tools for computational biology in Haskell continue to mature.
\section{What we think about it}

\subsection{Our Background}

\remark{I moved this section from the Intro as per NR's suggestion. ---AG}
All three authors are members of the Tufts University Department of Computer
Science.
Noah Daniels has taken a graduate seminar in advanced functional
programming, which included some Haskell, and spent ten years in industry as a
professional programmer in languages such as Ruby, C, and C++.
Andrew Gallant has taken a programming languages course heavily focused on 
functional programming, but with no Haskell.
Norman Ramsey is the local expert, but he wrote no code for this project.

\subsection{Barriers}
\remark{I feel like a lot of this is a summary of some of the problems we've
  discussed previously in the paper. Is that okay? A summary may be worth it. 
---AG}

As inexperienced functional programmers developing a complex piece of
software in Haskell, we encountered some barriers.
The first of these was profiling with GHC.
We previously mentioned that profiling with GHC required annotating
cost centers by hand, producing ugly code.
Beyond that, we also found it very difficult to profile code at all.
In particular, profiling required that all installed libraries be re-compiled
with profiling support,which was a process that we could not replicate
consistently.

Another barrier we faced was how to test our code using QuickCheck.
Indeed, the barrier was so formidable in our case that we still have not been 
able to use it.
The problems we faced specifically were two-fold: How do we start testing
a complex module? How can we use QuickCheck to generate random inputs with 
complex constraints? 
We wished to test invariants around beta-strand placement 
(potential solutions in our stochastic search), but found the QuickCheck generator
to be difficult to write.

Finally, one last barrier we faced was debugging runtime errors.
Runtime errors in the form of out-of-bounds indexing were particularly 
difficult to track down because they were not accompanied by a stack trace.
(While profiling with GHC could help remedy this, profiling was itself a 
barrier, as previously mentioned.)
We had to resort to polluting our code with calls to the \texttt{trace} function 
in order to track down where the out-of-bounds errors were occurring.

% - profiling 
% - debugging 
 % [ talk about profiling and debugging problems where?] 
% [ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]

\subsection{What you must know to succeed}
If others would like to follow in our footsteps, there are a few
things we believe one must know in order to succeed.
Firstly, some experience with functional programming concepts is a necessity. 
Specifically, coming from an imperative background, we found transforming input data 
from the HMMER file format without the ability to mutate state was difficult.
In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
and more natural than it would have been in C++.

Secondly, identifying existing tools and libraries from Hackage that are mature 
is also important.
We found that some packages that could have been immensely useful---like one 
that provides backtraces---were either not maintained or contained bugs.
Some other packages like Pads were functional but immature, which resulted in 
additional code to work around the parts of Pads that had not been implemented 
yet. 
In contrast, other packages such as \texttt{Data.Memocombinators} were
useful and mature.

Finally, a ``local expert'' or someone available to talk to with functional 
programming expertise, can make the transition to a language like Haskell 
easier.
We don't believe a local expert is absolutely necessary, but answers 
specifically related to functional programming idioms were easier to find when 
a local expert was available.

\subsection{Evaluation}
\remark{I think we need more in this section. ---AG}
In hindsight, we feel that there are several key facets of our project in 
Haskell that deserve evaluation.
Firstly, we found that existing tools and libraries were plentiful and 
typically helped us greatly.
In particular, \texttt{Data.Memocombinators} made dynamic 
programming with Viterbi exquisitely simple and elegant.
\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
Most of the tools we used came straight from Hackage, and although 
documentation of some form was almost always present, it was sometimes 
difficult to tell if a particular tool was right for us.
In particular, parts of the documentation were inaccessible to us as beginning 
functional programmers.
Documentation that was accessible to us typically contained straight-forward 
examples showing how to use that particular tool or library.

Another component worthy of evaluation is the debugging process.
Debugging in Haskell often takes the form of fixing type errors; which we think 
is a great thing.
However, in the case where debugging extends into runtime behavior, we couldn't 
find any clearly defined approach to effective debugging.
Stack traces are available via profiling, but this was difficult to find and 
as we mentioned, profiling had its own barriers.

Finally, we believe that we could have achieved a better division of modules in 
our project, but that conflicted with one our goals: to get publishable results
quickly. Despite being hasty programmers, we were still able to produce code 
that is easy to understand and easily extensible. Namely, in our experience, 
hasty Haskell code beats hasty code in a language like C or C++.

In summary, with Haskell, we didn't have to master a completely different 
design and development style but we did have to wrap our heads around 
functional programming.\remark{Andrew, look at the comments below this
part of the manuscript. See what you can do with them.}

% - What does the local expert do? Do you have to have one? 
%  
% - To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools). 
%  
% - No obvious body of knowledge on code improvement. 
%  
% - The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. 
% How to tell what's good? How did we discover what's good? 
% - Profiler can produce backtraces, but this was hard to discover.

% \appendix
% \section{Appendix Title}
% 
% This is the text of the appendix, if you need one.

\acks

We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford Larsen, and Nathan Ricci.

@@

\begingroup
\advance\leftskip by 0pt plus 10em
\emph{The Haskell wave passes, and if you slide off the crest, you
drown.}
\endgroup

>>>>

\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size

@@

\seclabel{awkward-profiling}

In two major areas, our lack of skills or knowledge resulted in ugly
Haskell.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, we found GHC's call-site and allocation
profiles did not provide enough information to be useful.
We~miss valgrind, whose callgrind profiler gives us enough information
without \emph{any} annotations in the source.

@@

\section{Our previous experience compared}

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
Our~prior experience therefore includes building,
maintaining, and extending research tools written in~C++.
One~such tool is~Matt,
a~structural protein aligner~\cite{Menke:2008wu}.
Matt comprises about 12,000 lines of~C++.
It uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.

We~modified Matt to use information about sequences as well as structure.
We~thought the modification would take three months, 
but it took most of a~year.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.
We~had hoped to add support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector
\cite{Menke:2010ti}.
The implementation took 12,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
For comparison, 
building the first functioning version of MRFy took three months of part-time work.
And we expect that higher-order functions will make extending MRFy
easy.
\remark{I'm thinking this \P\ is not so strong. ---NR}

Haskell encourages hasty programmers to slow down.
We~have to get the types right,
which makes it hard to write very large functions.
To~get the code to typecheck, we~have to write type signatures, which
also serve as documentation.
And once the types are accepted by the compiler,
it~is not much more work to write contracts for important functions.
MRFy is still hasty work.
Many types could be simplified;
we're~sure we've missed opportunities for abstraction;
and we know that MRFy's decomposition into modules could be improved.
But despite being hasty programmers, we produced code 
that is easy to understand and easy to extend.
Our hastily written Haskell beats
our hastily written Ruby and~C++.

When we look beyond our own group to computational biology more
broadly, our experiences with other software is better.
Little of it is written in functional languages, 
but much of the software shared by the community is excellent.
MRFy's training component was derived from that of~HMMER,
and
working with HMMER 
codebase was pleasant;
data structures and their
mutators are well documented. 
There is a 
BioHaskell library, part of which we use,
but it is not nearly as 
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that tools for computational biology in
Haskell continue to mature.

\section{What can you learn from our experience?}

If~you are a computational biologist, 
you can be productive in Haskell without extensive
experience.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a graduate seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell and other functional
languages for several medium-sized projects,
but his contribution was limited to minor ranting and refactoring.

\subsection{Barriers}

In~\secref{awkward-profiling} above, we~mention the awkwardness of the
cost-center annotations required for profiling.
But~we also found it difficult just to get the profiler working.
In~particular, profiling requires that all installed libraries be recompiled
with profiling support.
We~were not able to achieve such recompilations consistently.

We~faced another barrier in testing.
One of us (Ramsey) kept ranting that we should test our code using 
QuickCheck \cite{claessen:quickcheck},
but we could never figure out~how.
We~were stumped by two problems:
how to break down a complex module into parts that can usefully be
unit tested,
and how to implement QuickCheck generators for representations of complex
data like beta-strand placements.
We~know that QuickCheck is a powerful and proven tool,
yet we were not smart enough to figure out how to use~it.

Like other functional programmers, we~have found that often enough,
once we have the types right, the code is right.
But~much of our work involves computing with arrays and array indices,
and in that domain, types don't help much.
And we haven't been able to identify any systematic way of debugging
run-time errors in Haskell programs.
GHC's profiler can provide stack traces, but we~found this information
difficult to discover, and as noted above,
there are barriers to profiling.
We're~aware that debugging lazy functional programs has been a topic
of some research,\remark{Do we cite Hood, Hat, or Freya?}
but one of the biggest barriers we encountered to using Haskell is
that we have had to abandon our old approaches to debugging,
and we~see no clear path to follow.

\subsection{Information that will help you succeed}

If~you want to use Haskell in your research, we~believe that you must
have enough experience with functional programming that you can build
\emph{all} the code you need, not only the code that is easy to write
in a functional language.
Implementing Viterbi's equations in Haskell was pure joy.
Writing an iterative search in purely functional style was not
difficult.
When we had to transform data in the HMMER file format, \emph{without}
using mutable state the way the C++ code does, we~struggled.

You~also need to know that while the Haskell community offers many
enticing tools, libraries, and packages,
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.Memocombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell them from the wannabes
and the has-beens.

Most of the software we used came from Hackage, which is the Haskell
community's software repository.
We~sometimes found it hard to tell
if a particular Hackage package was right for us.
In particular, as beginning 
functional programmers, we~found some documentation inaccessible.
The documentation that we found most accessible typically contained
examples showing how to use a particular tool or library.
\remark{So what's our message here? What are we telling people to
think or do differently? ---NR}

As~in any endeavor, it's helpful to have access to experts.
We~would have been better off if our in-house expert had been an
enthusiastic student and not a busy professor.
But we have been surprised and pleased by the help available from
faraway experts on Stack Overflow and on Haskell mailing lists.
Although a local expert makes things easier, one is not
 absolutely necessary.

\section{Conclusion}

A~little knowledge of and a lot of love for functional programming was
enough for us to carry out a successful research project in a language
that is rarely used in computational biology.
If~you \emph{want} to use Haskell---or one of your graduate students
wants to use Haskell---you can
succeed.

% - No obvious body of knowledge on code improvement. 
%

\ifnotcutting
\acks

We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford
Larsen, and Nathan Ricci. 
\fi

@@

\begingroup
\parfillskip=0pt
\advance\leftskip by 0pt plus 30em
\emph{We live to surf the Haskell wave, but~if you slide off the crest, you
drown.}
\par
\endgroup

\break

\appendix

\section{Orphaned text}

In~our C++ work, the faults we encounter most frequently are
segmentation faults, 
memory errors caused by pointer arithmetic or uninitialized 
memory,
and heap exhaustion.
In~our Haskell work, the faults we encounter most frequently are
type errors and bounds-checking errors in array and vector accesses.
While we have found the bounds-checking errors difficult to debug,
they are less difficult than memory errors.
